{\TheTitle}\thanks{Submitted to the editors DATE. \funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}

### 2017-10-18 ssc-0-0.tex
revise the draft semi-implicit paper

For a system of equations~${dx\over dt}=f(x,t)$,
the semi-implicit formula with time step~$h$ is
\begin{equation}\label{eq:i}
    x_{n+1}=x_n+hf_{n+\frac12},
\end{equation}
where
\begin{equation}\label{eq:ii}
    f_{n+\frac12}=f\left({\textstyle\frac{x_n+x_{n+1}}{2}}\right).
\end{equation}
When this formula is applied to equations of a form similar to~\cref{eq:vi},
\begin{equation}\label{eq:iii}
	{dx\over dt}=-iHx,
\end{equation}
where~$H$ is a hermitian operator acting on the complex vector~$x$,
there is an advantage in principle because the norm of~$x$ is
conserved in the discretised solution.  {\bf Expand on this.}  However, the formula is
used mainly because it has been found to work well in practice.

% iterations
It is not obvious how to implement this formula, because~$x_{n+1}$
and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
usual way is a simple iteration.  To start,
$x_{n+1}$~is initialised to~$x_n$, and~$f_{n+\frac12}$ to~$f'(x_n)$.  Then each iteration
evaluates Equations~\cref{eq:i,eq:ii} in turn, and updates the values
of~$x_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the iterations generate a sequence of~$x_{n+1}^j$
and~$f_{n+\frac12}^j$, with
\begin{subequations}
\begin{align}
	\label{eq:iv} x_{n+1}^0&=x_n\qquad f_{n+\frac12}^0=f(x_n)\\
	\label{eq:v} x_{n+1}^{j+1}&=x_n+hf_{n+\frac12}^j
		\qquad f_{n+\frac12}^{j+1}=f\left({\textstyle\frac{x_n+x_{n+1}^j}{2}}\right),\qquad j\ge0.
\end{align}
\end{subequations}
This continues until the sequences converge, which is usually
achieved after a small, fixed number of iterations.  Any fixed point
of the iteration is a self-consistent solution to
Equations~\cref{eq:i,eq:ii}.  The remaining question is under what
conditions the iterations converge.

% steps stability
Generally, when a partial differential equation is discretised by
a given formula, the time step needs to be reduced in proportion
to the grid spacing in order for the discretised solution to remain
stable \cite{something}.  No one has analysed the stability of the
semi-implicit formula with iterations.  The stable time step has
had to be determined by trial and error, a time consuming process
for three-dimensional wave functions that are being simulated on
large grids.

This paper analyses the semi-implicit formula, in the case that it
is used to integrate linear Hamiltonian dynamics as in
Equation~\cref{eq:iii}.  The main result is that the stability
depends on a Nyquist-like condition, and the iterations converge
if and only if the largest eigenvalue of~$H$ is no more than~$2/h$.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant, the laplacian operator is usually
evaluated by a spectral method.  This is equivalent to expanding
the wave function~$\psi(q)$ over a basis of sinc wave packets
\begin{equation}
	\psi(q) = \sum x_i\phi_i(q),
\end{equation}
{\bf figure out range of $i$}
where
\begin{equation}
	\phi_i(q) = \frac1{\sqrt{l}} \sinc(\textstyle\frac{q}{l} - i).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc{q} = \frac{\sin(\pi q)}{\pi q},
\end{equation}
under which the functions~$\phi_i$ are an orthonormal basis for the
space of functions limited to bandwidth~$1/2l$.  The spectral method
reduces \cref{eq:vi} to an equation of the form~\cref{eq:iii} for
the coefficient vector~$x$, where
\begin{equation}
	H_{ij} = \int \phi_i(q) (-\nabla^2 + V(q)) \phi_j^\ast(q).
\end{equation}

At this point, the
iteration rules of Equation~\cref{eq:v} become
\begin{equation}
x_{n+1}^{j+1}=x_n+hf_{n+\frac12}^j\qquad f_{n+\frac12}^{j+1}
	=-{i\over2}Hx_n-{i\over2}Hx_{n+1}^j
\end{equation}
or
\begin{equation}
x_{n+1}^{j+1}=\left(1-{ih\over2}H\right)x_n-{ih\over2}Hx_{n+1}^j
\end{equation}
with solution {\bf check factor of 2}
\begin{equation}\label{eq:viii}
x_{n+1}^j=U_jx_n=\left(1-ihH-\frac{(hH)^2}2
	+{i(hH)^3\over 4}+\cdots+2\left({-ihH\over 2}\right)^j\right)x_n.
\end{equation}
The time step matrix~$U_j$ has the the same eigenvectors~$\omega_m$
as the hamiltonian matrix~$H$.  The eigenvalues of~$U_j$ are the
sums of a geometric series in~$\omega_m$, with the relation
between~$\omega_m$ and the sum~$\omega_m^j$ plotted in \cref{fig:ii}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_m$ of~$H$ and~$h\omega_m^j$ of~$U_j$.  The lightest line is for~$j=1$, and they get heavier for~$j=2$, 3, 4.  The exact time evolution operator~$U$ is unitary, with unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The sequence of~$U_j$ is a power series, which will converge if and
only if the angular frequencies~$\omega_m$ are bounded by the reciprocal time step~$2/h$, a
kind of Nyquist condition.  If the discretised hamiltonian has an
eigenstate~$x_\omega$ with a frequency~$\omega\gg 2/h$, then~$U$
will magnify the component of~$x_\omega$ in~$x$ by a factor close
to~$2(h\omega/2)^i$, and this component is a parasitic solution
growing steadily at the rate~$\frac{\ln2}h\frac{i}{h}\ln(\frac{h\omega}2)$.

{\bf Rewrite for clarity.}
The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{something}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

@@

\section*{Acknowledgments}
I would like to acknowledge the support of Swinburne University of Technology.

>>>>

% iterations
For a system of equations
\begin{equation}
	{ds\over dt}=f(s),
\end{equation}
the semi-implicit formula with time step~$h$ is
\begin{equation}\label{eq:i}
    s_{n+1}=s_n+hf_{n+\frac12},
	\qquad\text{where}\qquad 
    f_{n+\frac12}=f\left({\textstyle\frac{s_n+s_{n+1}}{2}}\right).
\end{equation}
It is not obvious how to implement this formula, because~$s_{n+1}$
and~$f_{n+\frac12}$ need to be evaluated self-consistently.  The
usual way is a simple iteration \cite{swx-5-12}.  To start, $s_{n+1}$~is initialised
to~$s_n$, and~$f_{n+\frac12}$ to~$f(s_n)$.  Then each iteration
evaluates Equation~\cref{eq:i}, and updates the
values of~$s_{n+1}$ and~$f_{n+\frac12}$.  More precisely, the
iterations generate a sequence of~$s_{n+1}^{(j)}$ and~$f_{n+\frac12}^{(j)}$,
with
\begin{subequations}
\begin{align}
	\label{eq:iv} s_{n+1}^{(0)}&=s_n,\qquad f_{n+\frac12}^{(0)}=f(s_n),\\
	\label{eq:v} s_{n+1}^{(j+1)}&=s_n+hf_{n+\frac12}^{(j)},
		\qquad f_{n+\frac12}^{(j+1)}=f\left({\textstyle\frac{s_n+s_{n+1}^{(j)}}{2}}\right)\qquad \text{when $j\ge0$}.
\end{align}
\end{subequations}
This continues until the sequences converge, which is usually
achieved after a small, fixed number of iterations.  Any fixed point
of the iteration satisfies
Equation~\cref{eq:i}.  The remaining question is under what
conditions the iterations converge.

% steps stability
Generally, when a partial differential equation is discretised by
a given formula, the time step needs to be reduced in proportion
to the grid spacing in order for the discretised solution to remain
stable \cite{2000-Trefethen-Spectral}.  No one has analysed the stability of the
semi-implicit formula with iterations.  The stable time step has
had to be determined by trial and error, a time consuming process
for three-dimensional wave functions that are being simulated on
large grids.

This paper analyses the semi-implicit formula, in the case that it
is used to integrate linear Hamiltonian dynamics as in
Equation~\cref{eq:iii}.  The main result is that the stability
depends on a Nyquist-like condition, and the iterations converge
if and only if the eigenvalues of the hamiltonian are bounded by the reciprocal time step~$2/h$.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant of it, the laplacian operator is usually
evaluated by a spectral method \cite{something}.  This is equivalent to expanding
the wave function~$\psi(x)$ over a basis of sinc wave packets
\begin{equation}
	\psi(x) = \sum_m s^m\phi_m(x),
\end{equation}
where, for a spatial grid of step~$l$,
\begin{equation}
	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc x = \frac{\sin\pi x}{\pi x},
\end{equation}
under which the set of functions~$\phi_m$ is an orthonormal basis for the
space of functions limited to spatial bandwidth~$1/2l$.  The spectral method
reduces \cref{eq:vi} to an equation for the complex coefficient vector~$s$ of the form
\begin{equation}\label{eq:iii}
	{ds\over dt}=-iHs,
\end{equation}
where~$H$ is a Hermitian matrix with elements
\begin{equation}
	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
\end{equation}

If the formula of Equation~\cref{eq:i} could be applied directly
to Equation~\cref{eq:iii}, the computed solution would be very
stable.  The coefficient vector~$s_n$ can be expanded over the
eigenvectors~$v_k$ of~$H$, as $s_n=\sum_k c^kv_k$.  \Cref{fig:iii}
shows the trajectory over a time step from~$s_n$ to~$s_{n+1}$,
projected onto the complex plane spanned by the eigenvector~$v_k$.
The exact solution that satisfies Equation~\cref{eq:iii}, with
initial condition~$s_n$, is~$s(h)=\sum_k c^ke^{-i\omega_kh}v_k$,
and its projection lies on the circle around the origin of
radius~$|c_k|$.

\begin{figure}[htbp]
  \centering
  \includegraphics{semii.1}
  \caption{A step of the semi-implicit formula, projected onto the
  complex plane spanned by the eigenvector~$v_k$ of~$H$.  The
  computed solution~$s_{n+1}$ lies on the same circle as the
  exact solution~$s(h)=\exp(-ihH)s_n$, but it rotates with angular frequency lower
  than~$\omega_k$.  See text for details.}
  \label{fig:iii}
\end{figure}

Suppose that the point~$s_{n+1}$ also lies on this circle.
Then~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
brought towards and away from the origin by moving~$s_{n+1}$ around
the circle, so there is some position for
which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, and this value of~$s_{n+1}$
satisfies Equation~\cref{eq:i}.  The only error introduced by the
formula is that~$c_k$ rotates with an angular frequency less
than~$\omega_k$, which is bounded by~$\pi/h$ as~$\omega_k$ becomes
large.

In practice, the iterations of~Equation~\cref{eq:vi} must be used
instead of~Equation~\cref{eq:i}, and this is less stable.  For
Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
iteration rule becomes
\begin{equation}
s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
\end{equation}
or
\begin{equation}
s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
\end{equation}
with solution
\begin{equation}\label{eq:viii}
s_{n+1}^{(j)}=U^{(j)}s_n
	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
\end{equation}
The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
in~$-ih\omega_k/2$, which agrees to second order in~$\omega_k$ with
the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$
plotted in \cref{fig:ii}, for the first four iterations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.  The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.  The exact time evolution operator~$U$ has unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The geometric series for~$U^{(j)}$ will converge if and only if the
angular frequencies~$\omega_k$ are bounded by the reciprocal time
step~$\frac2h$, a kind of Nyquist condition.  If the discretised
hamiltonian has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
will enlarge the component of~$v_k$ in~$s_n$ by a factor close
to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
thus a parasitic solution growing at the
rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.

{\bf Rewrite for clarity.}
The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{aop-70-67}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

@@

\section*{Acknowledgments}
I would like to acknowledge the support of Swinburne University of Technology.  Thanks to Run Yan Teh for suggestions that improved the presentation.
